---
title: "Conditional Probability"
subtitle: "Applied Statistics"
author: "MTH-361A | Spring 2025 | University of Portland"
date: "February 24, 2025"
output:
  slidy_presentation:
    font_adjustment: +5
    footer: "| MTH-361A Spring 2025 | <a href='../../index.html'>Back to the Course Website</a>"
    css: ../_style.css
bibliography: ../../references.bib
csl: ../../apa.csl
nocite: '@*'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r echo=FALSE, eval=TRUE,message=FALSE, warning=FALSE}
library(tidyverse)
library(openintro)
library(gghighlight)
library(latex2exp)
data(COL)
seed <- 42
```

## Objectives

:::: {.column width=15%}
::::

:::: {.column width=70%}
- **Introduce conditional probability**
- **Develop an understanding of dependent events**
- **Know how to compute conditional probabilities**
- **Activity: **
::::

:::: {.column width=15%}
::::

## Previously... (1/2)

**Basic Probability Rules**

| **Rule** | **Formula** |
|:---|:------|
| *Independence* | $P(A \cap B) = P(A)P(B)$ |
| *Joint (Union)* | $P(A \cup B) = P(A) + P(B) - P(A \cap B)$ |
| *Disjoint* | $P(A \cap B) = 0$ |
| *Complement* | If $P(A) + P(B) = 1$, then $1-P(A)=P(B)$. |

## Previously... (2/2)

| **Axiom** | **Statement** |
|:---|:------|
| $P(S) = 1$ | The sum of the probabilities for all outcomes in the sample space is equal to 1. |
| $P \in [0,1]$ | Probabilities are always positive and always between $0$ and $1$. |
| $P(A \cup B) = P(A) + P(B)$ | If events A and B are disjoint (mutually exclusive), then their probabilities can be added. |

## Sampling with Replacement



## Sampling without Replacement



## Drawing Cards



## Conditional Probability



## Balls in Urns Problem (Part 1/2)




## Bayes' Rule



## Balls in Urns Problem (Part 2/2)



## Activity: 



## References

::: {#refs}
:::
