---
title: "Inference for Linear Regression"
subtitle: "Applied Statistics"
author: "MTH-361A | Spring 2025 | University of Portland"
date: "April 7, 2025"
output:
  slidy_presentation:
    font_adjustment: +5
    footer: "| MTH-361A Spring 2025 | <a href='../../index.html'>Back to the Course Website</a>"
    css: ../_style.css
bibliography: ../../references.bib
csl: ../../apa.csl
nocite: '@*'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r echo=FALSE, eval=TRUE,message=FALSE, warning=FALSE}
library(tidyverse)
library(openintro)
library(gridExtra)
library(latex2exp)
library(kableExtra)
library(broom)
library(scales)
library(infer)
library(tidymodels)
data(COL)
seed <- 42
```

## Objectives

:::: {.column width=15%}
::::

:::: {.column width=70%}
- **Introduce inference for linear regression**
- **Develop an understanding of the sampling distribution of the slope and intercept**
- **Know how to compute the confidence intervals of the slope and intercept**
- **Activity: Determine Confidence Intervals of a Linear Model**
::::

:::: {.column width=15%}
::::

## Previously... (1/2)

**Linear Regression**

\[ y = \beta_0 + \beta_1 x + \epsilon  \]

* $y$ is the outcome, $x$ is the predictor, $\beta_0$ is the intercept, and $\beta_1$ is the slope. The notation $\epsilon$ is the model's error (residuals).
* We estimate the slope $b_1$ and intercept $b_0$ of the least squares regression line by minimizing the sum of squared residuals.
* The slope is given by $b_1 = \frac{s_y}{s_x} r$ , and the intercept is $b_0 = \bar{y} - b_1 \bar{x}$ , where $r$ is the correlation coefficient, $s_x$ and $s_y$ are the standard deviations of $x$ and $y$, and $\bar{x}$, $\bar{y}$ are their respective means.

## Previously... (2/2)

**Confidence Interval**

$$\text{point estimate} \pm z^* SE$$

where $SE$ is the standard error and $z^*$ is the critical z-value for a given confidence level assuming a standard normal distribution. Use $t_{df}^*$ for assuming a t-distribution given degrees of freedom $df$.

## Case Study I

```{r sandpop, fig.cap = "Revenue as a linear model of advertising dollars for a population of sandwich stores, in thousands of dollars.", out.width="70%", fig.align='center', fig.width=6,fig.height=4, warning=FALSE, message=FALSE}
set.seed(4747)
popsize <- 1000
ad <- rnorm(popsize, 4, 1)
rev <- 12 + 4.7 * ad + rnorm(popsize, 0, 8)
sandwich <- tibble(ad, rev)

ggplot(sandwich, aes(x = ad, y = rev)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = FALSE, fullrange = TRUE) +
  scale_x_continuous(labels = label_dollar(suffix = "K")) +
  scale_y_continuous(labels = label_dollar(suffix = "K")) +
  labs(
    x = "Amount spent on advertising",
    y = "Revenue",
    title = "Chain sandwich store",
    subtitle = "Hypothetical population"
  ) +
  coord_cartesian(ylim = c(0, 65)) + 
  theme_minimal()
```

## Case Study I: The Linear Model

The population model is: $$y_{revenue} = \beta_0 + \beta_1 x_{advertising} + e$$
where $y$ is the response, $x$ is the predictor, $\beta_0$ is the intercept, $\beta_1$ is the slope, and $e$ is the error term.

```{r echo=FALSE}
mod <- lm(rev ~ ad, data=sandwich)
```

The least squares regression model uses the data to find a best linear fit: 
$$\hat{y}_{revenue} = b_0 + b_1 x_{advertising}.$$
where $b_0 = 11.23$, $b_1 = 4.85$.

## Case Study I: Resdiual Analysis (1/2)

```{r echo=FALSE}
preds <- sandwich %>% 
  mutate(rev_hat = predict(mod, sandwich),
         resids = rev-rev_hat)
```

```{r echo=FALSE, fig.align='center', fig.height=2,fig.width=6, out.width="80%"}
p1 <- ggplot(preds,aes(x=resids)) + 
  geom_histogram(bins=30) + 
  xlab("residuals") + 
  ylab("count") + 
  theme_minimal()

p2 <- ggplot(preds,aes(x=rev_hat,y=resids)) + 
  geom_point() + 
  geom_hline(yintercept = 0, color="blue", linetype="dashed") + 
  xlab("predicted revenue") + 
  ylab("residuals") + 
  theme_minimal()

grid.arrange(p1,p2,nrow=1)
```

* The residuals appear approximately normal, supporting the normality assumption.
* The residual plot shows roughly constant variance, suggesting homoskedasticity.
* Slight clustering near the center may require further investigation but otherwise the residuals are independent of the model (no pattern).

## Case Study I: Resdiual Analysis (2/2)

A **Q-Q plot (quantile-quantile plot)** is a graphical tool used to assess whether a dataset follows a particular theoretical distribution—most commonly, the normal distribution.

:::: {.column width=49%}
```{r echo=FALSE, fig.align='center', fig.height=3, fig.width=4, out.width='100%'}
ggplot(preds,aes(sample=resids)) + 
  stat_qq() + 
  stat_qq_line(color="blue") + 
  xlab("theoretical quantiles") + 
  ylab("sample quantiles") + 
  theme_minimal()
```
::::

:::: {.column width=50%}
* In regression, we often use a Q-Q plot of residuals to check if they are normally distributed --one of the important assumptions for valid inference for linear regression.
* Quick interpretation:
  - Straight line means normality is a reasonable assumption.
  - Curved or S-shaped means deviations from normality.
  - Outliers far from the line means possible extreme values.
::::

## Case Study I: Variability of the Statistic (1/4)

```{r echo = FALSE,warning=FALSE, message=FALSE}
set.seed(470)
sandwich2 <- sandwich %>%
  sample_n(size = 20)
sandwich3 <- sandwich %>%
  sample_n(size = 20)
sandwich_many <- sandwich %>%
  rep_sample_n(size = 20, replace = FALSE, reps = 100)
```

```{r fig.cap = "A random sample of 20 stores from the entire population. A linear trend between advertising and revenue continues to be observed.", out.width="70%", fig.align='center', fig.width=6,fig.height=4, warning=FALSE, message=FALSE}
ggplot(sandwich2, aes(x = ad, y = rev)) +
  geom_point(size = 3, fill = IMSCOL["green", "full"], color = "#FFFFFF", shape = 22) +
  geom_smooth(method = "lm", se = FALSE, fullrange = TRUE, color = IMSCOL["green", "full"]) +
  scale_x_continuous(labels = label_dollar(suffix = "K")) +
  scale_y_continuous(labels = label_dollar(suffix = "K")) +
  labs(
    x = "Amount spent on advertising",
    y = "Revenue",
    title = "Chain sandwich store",
    subtitle = "Random sample of 20 stores"
  ) +
  coord_cartesian(ylim = c(0, 65)) + 
  theme_minimal()
```
  
## Case Study I: Variability of the Statistic (2/4)

A second sample of size 20 also shows a positive trend!

```{r fig.cap = "A different random sample of 20 stores from the entire population. Again, a linear trend between advertising and revenue is observed.", out.width="70%", fig.align='center', fig.width=6,fig.height=4, warning=FALSE, message=FALSE}

ggplot(sandwich3, aes(x = ad, y = rev)) +
  geom_point(size = 3, fill = IMSCOL["gray", "full"], color = "#FFFFFF", shape = 23) +
  geom_smooth(method = "lm", se = FALSE, fullrange = TRUE, color = IMSCOL["gray", "full"]) +
  scale_x_continuous(labels = label_dollar(suffix = "K")) +
  scale_y_continuous(labels = label_dollar(suffix = "K")) +
  labs(
    x = "Amount spent on advertising",
    y = "Revenue",
    title = "Chain sandwich store",
    subtitle = "Random sample of 20 stores"
  ) +
  coord_cartesian(ylim = c(0, 65)) + 
  theme_minimal()
```

## Case Study I: Variability of the Statistic (3/4)

```{r fig.cap = "The linear models from the two different random samples are quite similar, but they are not the same line.", out.width="70%", fig.align='center', warning=FALSE, message=FALSE, fig.width=6,fig.height=4}
ggplot() +
  geom_point(data = sandwich2, aes(x = ad, y = rev), 
             size = 3, , shape = 22,
             fill = IMSCOL["green", "full"], color = "#FFFFFF") +
  geom_smooth(data = sandwich2, aes(x = ad, y = rev), 
              method = "lm", se = FALSE, fullrange = TRUE, 
              color = IMSCOL["green", "full"]) +
  geom_point(data = sandwich3, aes(x = ad, y = rev), 
             size = 3, , shape = 23,
             fill = IMSCOL["gray", "full"], color = "#FFFFFF") +
  geom_smooth(data = sandwich3, aes(x = ad, y = rev), 
              method = "lm", se = FALSE, fullrange = TRUE, 
              color = IMSCOL["gray", "full"]) +
  scale_x_continuous(labels = label_dollar(suffix = "K")) +
  scale_y_continuous(labels = label_dollar(suffix = "K")) +
  labs(
    x = "Amount spent on advertising",
    y = "Revenue",
    title = "Chain sandwich store",
    subtitle = "Two random samples of 20 stores"
  ) +
  coord_cartesian(ylim = c(0, 65)) + 
  theme_minimal()
```

## Case Study I: Variability of the Statistic (4/4)

```{r slopes, fig.cap = "If repeated samples of size 20 are taken from the entire population, each linear model will be slightly different. The red line provides the linear fit to the entire population.", out.width="70%", fig.align='center', fig.width=6,fig.height=4, warning=FALSE, message=FALSE}
ggplot() +
  geom_smooth(
    data = sandwich_many, aes(x = ad, y = rev, group = replicate),
    method = "lm", se = FALSE, fullrange = TRUE, 
    color = IMSCOL["gray", "f2"], size = 0.5
  ) +
  geom_smooth(
    data = sandwich, aes(x = ad, y = rev), method = "lm", se = FALSE,
    fullrange = TRUE, color = IMSCOL["red", "full"]
  ) +
  scale_x_continuous(labels = label_dollar(suffix = "K")) +
  scale_y_continuous(labels = label_dollar(suffix = "K")) +
  labs(
    x = "Amount spent on advertising",
    y = "Revenue",
    title = "Chain sandwich store",
    subtitle = "100 random samples of 20 stores"
  ) +
  coord_cartesian(ylim = c(0, 65)) + 
  theme_minimal()
```
  
## Case Study I: Sampling Distribution of the Slope Estimate

```{r sand20lm, fig.cap="Variability of slope estimates taken from many different samples of stores, each of size 20.", out.width="70%", warning=FALSE, message=FALSE, fig.align='center', fig.width=6,fig.height=4}
sandwich_many_lm <- sandwich_many %>%
  group_by(replicate) %>%
  do(tidy(lm(rev ~ ad, data = .))) %>%
  filter(term == "ad")

ggplot(sandwich_many_lm, aes(x = estimate)) +
  geom_histogram(binwidth = 0.5) +
  labs(
    x = "Slope estimate",
    y = "Count",
    title = "Chain sandwich store",
    subtitle = "100 random samples of 20 stores"
    ) + 
  geom_vline(xintercept = mean(sandwich_many_lm$estimate),color="red") + 
  annotate(x=mean(sandwich_many_lm$estimate),y=+Inf,label="mean slope",vjust=11,geom="label",color="red") +
  geom_vline(xintercept = 4.8,color="blue") + 
  annotate(x=4.8,y=+Inf,label="actual slope",vjust=13,geom="label",color="blue") +
  theme_minimal()
```

## Case Study I: Confidence Interval (1/2)

The 95% Confidence interval for the slope and intercept using the bootstrapped samples. Here, we are using the percentiles of the bootstrapped samples to estimate the confidence interval.

```{r message=FALSE, warning=FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.width=6,fig.height=3, out.width="60%"}
boots <- bootstraps(sandwich, times = 100, apparent = TRUE)

fit_lm_on_bootstrap <- function(split) {
    lm(rev ~ ad, analysis(split))
}

boot_models <-
  boots %>% 
  dplyr::mutate(model = map(splits, fit_lm_on_bootstrap),
         coef_info = map(model, tidy))

boot_coefs <- 
  boot_models %>% 
  unnest(coef_info)

percentile_intervals <- int_pctl(boot_models, coef_info)
percentile_intervals %>% 
  select(term,.lower,.estimate,.upper)

ggplot(boot_coefs, aes(estimate)) +
  geom_histogram(bins = 30) +
  facet_wrap( ~ term, scales = "free") +
  labs(title="mtcars", subtitle = "mpg ~ wt") +
  labs(caption = "95% confidence interval parameter estimates, intercept + estimate") +
  geom_vline(aes(xintercept = .lower), data = percentile_intervals, col = '#528fad') +
  geom_vline(aes(xintercept = .upper), data = percentile_intervals, col = '#528fad') + 
  theme_minimal()
```

## Case Study I: Confidence Interval (2/2)

```{r slopes2, fig.cap = "If repeated samples of size 20 are taken from the entire population, each linear model will be slightly different. The red line provides the linear fit to the entire population.", out.width="70%", fig.align='center', fig.width=6,fig.height=4, warning=FALSE, message=FALSE}
ggplot() +
  geom_smooth(
    data = sandwich_many, aes(x = ad, y = rev, group = replicate),
    method = "lm", se = FALSE, fullrange = TRUE, 
    color = IMSCOL["gray", "f2"], size = 0.5
  ) +
  geom_smooth(
    data = sandwich, aes(x = ad, y = rev), method = "lm", se = FALSE,
    fullrange = TRUE, color = IMSCOL["red", "full"]
  ) +
  geom_abline(intercept = 9.834915	, slope = 5.236326) +
  geom_abline(intercept = 13.218205	, slope = 4.358568) +
  scale_x_continuous(labels = label_dollar(suffix = "K")) +
  scale_y_continuous(labels = label_dollar(suffix = "K")) +
  labs(
    x = "Amount spent on advertising",
    y = "Revenue",
    title = "Chain sandwich store",
    subtitle = "100 random samples of 20 stores"
  ) +
  coord_cartesian(ylim = c(0, 65)) + 
  theme_minimal()
```

## Case Study I: Best Fit Line and Confidence Interval

```{r sandpop2, fig.cap = "Revenue as a linear model of advertising dollars for a population of sandwich stores, in thousands of dollars.", out.width="70%", fig.align='center', fig.width=6,fig.height=4, warning=FALSE, message=FALSE}
set.seed(4747)
popsize <- 1000
ad <- rnorm(popsize, 4, 1)
rev <- 12 + 4.7 * ad + rnorm(popsize, 0, 8)
sandwich <- tibble(ad, rev)

ggplot(sandwich, aes(x = ad, y = rev)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = TRUE, fullrange = TRUE) +
  scale_x_continuous(labels = label_dollar(suffix = "K")) +
  scale_y_continuous(labels = label_dollar(suffix = "K")) +
  labs(
    x = "Amount spent on advertising",
    y = "Revenue",
    title = "Chain sandwich store",
    subtitle = "Hypothetical population"
  ) +
  coord_cartesian(ylim = c(0, 65)) + 
  theme_minimal()
```

We can conclude that we are 95% confident that the true slope that models the spending amount on advertising versus revenue is 4.36 to 5.23. Note that the slope of 0 --which means no association between the variables-- is not within the confidence interval, suggesting a statistically significant association.

## Activity: Determine Confidence Intervals of a Linear Model

1. Log-in to Posit Cloud and open the R Studio assignment *M 4/7 - Determine Confidence Intervals of Linear Model*.
2. Make sure you are in the current working directory. Rename the `.Rmd` file by replacing `[name]` with your name using the format `[First name][Last initial]`. Then, open the `.Rmd` file.
3. Change the author in the YAML header.
4. Read the provided instructions.
5. Answer all exercise problems on the designated sections.

## References

::: {#refs}
:::
