---
title: "Selecting Variables for Multiple Linear Regression"
subtitle: "Applied Statistics"
author: "[name]"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Syllabus Reminders

- You must submit your worksheet individually by end-of-class or end-of-day. Your name must exist in your worksheet and the names of your collaborators.
- Worksheets are marked mostly on completion, and partially on correctness. It will be marked either pass or fail, there will no detailed feedback on worksheets, and no opportunities for revisions and make-up.

# Instructions

- For the exercises, please provide your answers in full sentences.
- Write your answers by replacing the text "[Write your answer here]".
- Whenever a problem asks you to run something in R, create an R code chunk and write your R commands inside that chunk.
- If a problem requires modifying an existing R code chunk, make the necessary changes at the appropriate lines within the given code.
- Please read the problems fully and carefully before answering.
- When finished, knit this `.Rmd` to `.html`.
- Submit this `.Rmd` and the recently knitted `.html` to Moodle.

# Packages

The following R packages have already been loaded for you, so there's no need to load them yourself—they will be used in the problem sets. If you need any additional packages, you can load them here as well.

```{r message=FALSE, warning=FALSE}
# load packages
library(tidyverse)
library(openintro)
library(statsr)
library(broom)
library(gridExtra)
library(MASS)
```

# Collaborators

::: {style="color: blue;"}
[Write your answer here]
:::

# Exercises

## Problem 1
 
In this worksheet you will explore the `mariokart` data set from the `openintro` package. Use `?mariokart` to determine the variables of the data set.

a. How many observations are there and what does each represent?
b. How many numerical and categorical variables in the data?
c. Subset the data by selecting only the numerical variables.

::: {style="color: blue;"}
[Write your answer here]
:::

## Problem 2

Continues from Problem (1). Explore the data subset by visualizing the distribution of the variables and their relationships to the `total_pr` variable and to each other.

a. Are there any co-linear variables?
b. If so, do you need to transform it into log-scale?

::: {style="color: blue;"}
[Write your answer here]
:::

## Problem 3

Use the `lm()` function to estimate the coefficients of the multiple linear regression of the numerical variables, where the variable `total_pr` is the response.

a. Write the equation of the model.
b. What are the p-values of each estimate and what does it mean?

::: {style="color: blue;"}
[Write your answer here]
:::

## Problem 4

Repeat the Problems (2) and (3), but using the categorical variables. How does this result compare to just using the numerical variables.

::: {style="color: blue;"}
[Write your answer here]
:::

## Problem 5

Repeat the Problems (2) and (3), but using all numerical and categorical variables. How does this result compare to just using the numerical variables.

::: {style="color: blue;"}
[Write your answer here]
:::
