---
title: "Discrete Random Variabes & <br>Probability Mass Functions"
subtitle: "Applied Statistics"
author: "MTH-361A | Spring 2025 | University of Portland"
date: "February 12, 2025"
output:
  slidy_presentation:
    font_adjustment: +5
    footer: "| MTH-361A Spring 2025 | <a href='../../index.html'>Back to the Course Website</a>"
    css: ../_style.css
bibliography: ../../references.bib
csl: ../../apa.csl
nocite: '@*'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r message=FALSE, warning=FALSE}
library(gridExtra)
```

## Objectives

:::: {.column width=15%}
::::

:::: {.column width=70%}
- **Develop an understanding of sampling with or without replacement**
- **Introduce conditional probability**
- **Know how binomial and geometric random variables work**
- **Activity: Compute Empirical vs Theoretical Probabilities**
::::

:::: {.column width=15%}
::::

## Previously... (1/2)

**Random Variables (R.V.)**

A **random variable** is a numerical outcome of a random experiment. It assigns a number to each possible outcome in a sample space.

In other words, a random variable is **a function that maps the sample space into real numbers**.

**Types**

* *Discrete Random Variable:* Takes on a countable number of values.
* *Continuous Random Variable:* Takes on any value in an interval.

## Previously... (2/2)

**The Bernoulli R.V.**

A **Bernoulli r.v.** represents a single experiment with two possible outcomes: success (1) with probability $p$ and failure (0) with probability $1-p$.

$$X \sim \text{Bern}(p)$$

The **Bernoulli distribution** is a **probability function** that computes the probability of the Bernoulli random variable.

The **probability mass function** of the *Bernoulli r.v.* is given by $$P(X = x) = p^x (1-p)^{1-x}, \ \ x \in \{0,1\}$$ where $p$ is the parameter (also the probability of "success").

## Sampling with Replacement

The **sampling with replacement** is a sampling method where each selected item is returned to the population before the next selection.

:::: {.column width=50%}
**Key Characteristics:**

* *Independent Trials:* Each draw does not affect the next.
* *Same Probability:* Every item has the same chance of being selected each time.
* *Allows Repetition:* The same item can be chosen multiple times.
::::

:::: {.column width=49%}
**Example:**

* Drawing a card from a deck, recording its value, and putting it back before drawing again.
* Rolling a die multiple times, where each roll does not influence the next.
::::

## Sampling without Replacement

The **sampling without replacement** is a sampling method where each selected item is not returned to the population before the next selection.

:::: {.column width=50%}
**Key Characteristics:**

* *Dependent Trials:* Each draw changes the population.
* *Changing Probability:* The chance of selecting an item increases or decreases based on previous
* *No Repetition:* Once an item is chosen, it cannot be selected again.
::::

:::: {.column width=49%}
**Example:**

* Drawing a card from a deck without returning it, meaning the next draw is from a smaller set.
* Drawing items from jars without return it before drawing the next item.
::::

## Multiple Bernoulli Trials

**Bernoulli R.V.**

$$X \sim \text{Bern}(p)$$

**Probability Mass Function (PMF)**

$$P(X = x) = p^x (1-p)^{1-x}, \ \ x \in \{0,1\}$$

**Multiple bernoulli trials** is a sequence of independent Bernoulli trials, each with the same probability *p*.

* **Key assumption:** Trials are independent (i.e., one outcome does not affect the next).

* Leads to important distributions:
    - Geometric Distribution (number of trials until the first success)
    - Binomial Distribution (sum of multiple Bernoulli trials)

## The Geometric R.V.

A **geometric R.V.** is a discrete random variable that represents the number of Bernoulli trials until the first success. Each trial is independent, with a constant probability of success $p$.

$$X \sim \text{Geom}(p)$$

:::: {.column width=50%}
**Possible Outcomes:**

\[
\begin{aligned}
1 & \longrightarrow 0 \text{ failure until success} \\
0,1 & \longrightarrow 1 \text{ failure until success} \\
0,0,1 & \longrightarrow 2 \text{ failures until success} \\
0,0,0,1 & \longrightarrow 3 \text{ failures until success} \\
 & \vdots \\
0,0,0,\cdots,1 & \longrightarrow k \text{ failures until success} \\
\end{aligned}
\]
::::

:::: {.column width=49%}
**Probabilities:**

\[
\begin{aligned}
1 & \longrightarrow p \\
0,1 & \longrightarrow (1-p) p \\
0,0,1 & \longrightarrow (1-p)^2 p \\
0,0,0,1 & \longrightarrow (1-p)^3 p \\
 & \vdots \\
0,0,0,\cdots,1 & \longrightarrow (1-p)^k p \\
\end{aligned}
\]
::::

## The Geometric R.V.: PMF

The geometric R.V. $X \sim \text{Geom}(p)$ has infinite possible outcomes (or infinite sized sample space) with PMF given by $$P(X=k) = (1-p)^k p$$ where $k = 1,2,3, \cdots$ and $p$ is the probability of success.

**Example:**

* What is the probability of success after 5 trials with $p=0.50$?

$$P(X=5) = (1-0.50)^5 0.50 \approx 0.016$$

## The Binomial R.V.

A **binomial R.V.** is a discrete random variable representing the number of successes in $n$ independent Bernoulli trials, each with probability of success $p$.

$$X \sim \text{Binom}(n,p)$$

:::: {.column width=50%}
**Possible Outcomes:**

Suppose $n = 3$.

\[
\begin{aligned}
0,0,0 & \longrightarrow 3 \text{ failure and } 0 \text{ success} \\
0,0,1 & \longrightarrow 2 \text{ failure and } 1 \text{ success} \\
0,1,0 & \longrightarrow 2 \text{ failure and } 1 \text{ success} \\
0,1,1 & \longrightarrow 1 \text{ failure and } 2 \text{ success} \\
1,0,0 & \longrightarrow 2 \text{ failure and } 1 \text{ success} \\
1,0,1 & \longrightarrow 1 \text{ failure and } 2 \text{ success} \\
1,1,0 & \longrightarrow 1 \text{ failure and } 2 \text{ success} \\
1,1,1 & \longrightarrow 0 \text{ failure and } 3 \text{ success} \\
\end{aligned}
\]
::::

:::: {.column width=49%}
**Probabilities:**

Suppose $n = 3$.
\[
\begin{aligned}
0,0,0 & \longrightarrow (1-p)^3 \\
0,0,1 & \longrightarrow (1-p)^2 p \\
0,1,0 & \longrightarrow (1-p)^2 p \\
1,0,0 & \longrightarrow (1-p)^2 p \\
0,1,1 & \longrightarrow (1-p) p^2 \\
1,0,1 & \longrightarrow (1-p) p^2 \\
1,1,0 & \longrightarrow (1-p) p^2 \\
1,1,1 & \longrightarrow p^3
\end{aligned}
\]
::::

## The Binomial R.V.: PMF

The binomial R.V. $X \sim \text{Binom}(n,p)$ has finite possible outcomes with PMF given by $$P(X=k) = \binom{n}{k} p^k (1-p)^{n-k}$$ where $k = 1,2,3, \cdots, n$, $p$ is the probability of success, and $\binom{n}{k} = \frac{n!}{k! (n-k)!}$ is the binomial coefficient

**Example:**

* What is the probability of getting 4 successes in 10 trials with $p=0.50$?

$$P(X=4) =\binom{10}{4} (0.50)^4 (1-0.50)^{n-k} \approx 0.205$$

## Activity: Compute Empirical vs Theoretical Probabilities

1. Make sure you have a copy of the *W 2/12 Worksheet*. This will be handed out physically and it is also digitally available on Moodle.
2. Work on your worksheet by yourself for 10 minutes. Please read the instructions carefully. Ask questions if anything need clarifications.
3. Get together with another student.
4. Discuss your results.
5. Submit your worksheet on Moodle as a `.pdf` file.

## References

::: {#refs}
:::
