---
title: "Inference for Linear Regression"
subtitle: "Applied Statistics"
author: "MTH-361A | Spring 2025 | University of Portland"
date: "April 4, 2025"
output:
  slidy_presentation:
    font_adjustment: +5
    footer: "| MTH-361A Spring 2025 | <a href='../../index.html'>Back to the Course Website</a>"
    css: ../_style.css
bibliography: ../../references.bib
csl: ../../apa.csl
nocite: '@*'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r echo=FALSE, eval=TRUE,message=FALSE, warning=FALSE}
library(tidyverse)
library(openintro)
library(gridExtra)
library(latex2exp)
library(kableExtra)
library(broom)
data(COL)
seed <- 42
```

## Objectives

:::: {.column width=15%}
::::

:::: {.column width=70%}
- 
- **Activity: **
::::

:::: {.column width=15%}
::::

## Previously... 



## Activity: 

1. Make sure you have a copy of the *F 4/4 Worksheet*. This will be handed out physically and it is also digitally available on Moodle.
2. Work on your worksheet by yourself for 10 minutes. Please read the instructions carefully. Ask questions if anything need clarifications.
3. Get together with another student.
4. Discuss your results.
5. Submit your worksheet on Moodle as a `.pdf` file.

## References

::: {#refs}
:::
