---
title: "Simulate Multiple Bernoulli Trials"
subtitle: "Applied Statistics"
author: "<span style='color:blue'>Solutions</span>"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Syllabus Reminders

- You must submit your worksheet individually by end-of-class or end-of-day. Your name must exist in your worksheet and the names of your collaborators.
- Worksheets are marked mostly on completion, and partially on correctness. It will be marked either pass or fail, there will no detailed feedback on worksheets, and no opportunities for revisions and make-up.

# Instructions

- For the exercises, please provide your answers in full sentences.
- Write your answers by replacing the text "[Write your answer here]".
- Whenever a problem asks you to run something in R, create an R code chunk and write your R commands inside that chunk.
- If a problem requires modifying an existing R code chunk, make the necessary changes at the appropriate lines within the given code.
- Please read the problems fully and carefully before answering.
- When finished, knit this `.Rmd` to `.html`.
- Submit this `.Rmd` and the recently knitted `.html` to Moodle.

# Packages

The following R packages have already been loaded for you, so there's no need to load them yourself—they will be used in the problem sets. If you need any additional packages, you can load them here as well.

```{r message=FALSE, warning=FALSE}
# load packages
library(tidyverse)
```

# Collaborators

::: {style="color: blue;"}
<span style='color:blue'>Solutions</span>
:::

# Exercises

## Problem 1

The following code simulates a sequence of Bernoulli trials, where each trial follows a Bernoulli distribution with a "success" probability of $p=0.60$ and a "failure" probability of $1-p=0.40$.

```{r eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
# set hyperparameters
set.seed(42) # for reproducibility
n <- 10 # number of trials

# set parameter and PMF of the Bernoulli r.v
X <- c("0","1") # outcomes ("0"="failure","1"="success")
p <- 0.6 # probability of success
bern_pmf <- c(1-p,p)

# simulate n Bernoulli trials
samples <- sample(X,size=n,prob=bern_pmf,replace=TRUE)

# convert samples into tibble form and compute proportions
samples_tib <- tibble(X=samples) %>% 
  group_by(X) %>% 
  summarise(count=n(),proportion=count/n)

# view tibble
samples_tib
```

Modify the above code sequence to simulate Bernoulli trials with 3 different "success" probabilities:

  a. $p_1 = 0.25$
  b. $p_2 = 0.50$
  c. $p_3 = 0.75$

::: {style="color: blue;"}
```{r eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
# set hyperparameters
set.seed(42) # for reproducibility
n <- 1000 # number of trials

# set parameter and PMF of the Bernoulli r.v
X <- c("0","1") # outcomes ("0"="failure","1"="success")
p_a <- 0.25 # part (a) p
p_b <- 0.50 # part (b) p
p_c <- 0.75 # part (c) p
bern_pmf_a <- c(1-p_a,p_a) # part (a) pmf
bern_pmf_b <- c(1-p_b,p_b) # part (b) pmf
bern_pmf_c <- c(1-p_c,p_c) # part (b) pmf

# simulate n Bernoulli trials
samples_a <- sample(X,size=n,prob=bern_pmf_a,replace=TRUE) # part (a) samples
samples_b <- sample(X,size=n,prob=bern_pmf_b,replace=TRUE) # part (b) samples
samples_c <- sample(X,size=n,prob=bern_pmf_c,replace=TRUE) # part (c) samples

# convert samples into tibble form and compute proportions
samples_tib_a <- tibble(X=samples_a) %>% # part (a) samples as tibble
  group_by(X) %>% 
  summarise(count=n(),proportion=count/n)
samples_tib_b <- tibble(X=samples_b) %>% # part (b) samples as tibble
  group_by(X) %>% 
  summarise(count=n(),proportion=count/n)
samples_tib_c <- tibble(X=samples_c) %>% # part (c) samples as tibble
  group_by(X) %>% 
  summarise(count=n(),proportion=count/n)

# view tibble
samples_tib_a
samples_tib_b
samples_tib_c
```
:::

## Problem 2

Describe your observation of the proportion of outcomes ("success" and "failure") as you increase the number of trials for each $p$ in Problem (1).

::: {style="color: blue;"}
As the number of trials increase, the "success" and "failure" proportions of the samples are getting closer to their respective probabilities.
:::
