---
title: "Conduct a Hypothesis Test for Linear Regression"
subtitle: "Applied Statistics"
author: "[name]"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Syllabus Reminders

- You must submit your worksheet individually by end-of-class or end-of-day. Your name must exist in your worksheet and the names of your collaborators.
- Worksheets are marked mostly on completion, and partially on correctness. It will be marked either pass or fail, there will no detailed feedback on worksheets, and no opportunities for revisions and make-up.

# Instructions

- For the exercises, please provide your answers in full sentences.
- Write your answers by replacing the text "[Write your answer here]".
- Whenever a problem asks you to run something in R, create an R code chunk and write your R commands inside that chunk.
- If a problem requires modifying an existing R code chunk, make the necessary changes at the appropriate lines within the given code.
- Please read the problems fully and carefully before answering.
- When finished, knit this `.Rmd` to `.html`.
- Submit this `.Rmd` and the recently knitted `.html` to Moodle.

# Packages

The following R packages have already been loaded for you, so there's no need to load them yourself—they will be used in the problem sets. If you need any additional packages, you can load them here as well.

```{r message=FALSE, warning=FALSE}
# load packages
library(tidyverse)
library(openintro)
library(statsr)
library(broom)
library(gridExtra)
library(MASS)
```

# Collaborators

::: {style="color: blue;"}
[Write your answer here]
:::

# Exercises

## Problem 1

In this worksheet you will be using the `murders` data set from the `MASS` package. Use `?murders` to determine the type of variables are in this data set, the number of observations, and what are the observations represent.

::: {style="color: blue;"}
[Write your answer here]
:::

## Problem 2

The goal is to model the murders per mile based on the poverty level. 

a. Create a scatter plot that shows the two numerical variables of interest.
b. Fit a linear regression line using the `lm()` function, and write the linear equation for the population model and the estimated linear model.
c. Interpret the slope of the best fit line.

::: {style="color: blue;"}
[Write your answer here]
:::

## Problem 3

Assess the linear model determined in Problem (2) by analysing the residuals. You can compute the residuals using the `predict()` function. Are the assumptions for inference for linear regression satisified?

::: {style="color: blue;"}
[Write your answer here]
:::

## Problem 4

What are the hypotheses for evaluating whether the slope of the model predicting annual murder rate from poverty percentage is different than 0?

::: {style="color: blue;"}
[Write your answer here]
:::

## Problem 5

View the result of the linear model determined in Problem (2) by using the `summary()` function.

a. State the conclusion of the hypothesis test in context of the data.
b. What does this say about whether poverty percentage is a useful predictor of annual murder rate?

::: {style="color: blue;"}
[Write your answer here]
:::
