---
title: "Chi-Squared Tests"
subtitle: "Applied Statistics"
author: "MTH-361A | Spring 2025 | University of Portland"
date: "March 21, 2025"
output:
  slidy_presentation:
    font_adjustment: +5
    footer: "| MTH-361A Spring 2025 | <a href='../../index.html'>Back to the Course Website</a>"
    css: ../_style.css
bibliography: ../../references.bib
csl: ../../apa.csl
nocite: '@*'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r echo=FALSE, eval=TRUE,message=FALSE, warning=FALSE}
library(tidyverse)
library(openintro)
library(gridExtra)
library(latex2exp)
data(COL)
seed <- 42
```

## Objectives

:::: {.column width=15%}
::::

:::: {.column width=70%}
- **Develop an understanding of using the chi-squared distribution**
- **Know how to conduct and interpret chi-squared test results**
- **Activity: **
::::

:::: {.column width=15%}
::::

## Previously... (1/3)



## Previously... (2/3)


## Previously... (3/3)



## Activity: 

1. Make sure you have a copy of the *F 3/21 Worksheet*. This will be handed out physically and it is also digitally available on Moodle.
2. Work on your worksheet by yourself for 10 minutes. Please read the instructions carefully. Ask questions if anything need clarifications.
3. Get together with another student.
4. Discuss your results.
5. Submit your worksheet on Moodle as a `.pdf` file.

## References

::: {#refs}
:::
